A Short and Mechanized Consistency Proof for a Dependent Type Theory

### 2023-12-18 paper/paper.tex
Add proof about coherence

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$. % The $*$ symbol is more commonly
% used for the transitive and reflexive closure of a relation. However,
% since we can show that parallel reduction is reflexive, the transitive
% closure of parallel reduction must also be reflexive.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation, and in fact, we
will use coherence as the equational theory for our type conversion rule.

The following lemma shows us that parallel reduction is
reflexive. As a result, we only need to take its transitive closure
for the resulting relation to also be reflexive.
\begin{lemma}[Par Refl]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}

@@

A more conventional presentation would instead use full beta reduction
instead of parallel reduction as the base for the definition of
coherence. However, the diamond

>>>>

Figure~\ref{fig:par} shows the definition of the parallel reduction
relation, which takes the form $[[a => b]]$. We use $[[a =>+ b]]$ to
represent its transitive closure, which in turn allows us to define
the coherence relation $[[a <=> b]]$. % The $*$ symbol is more commonly
% used for the transitive and reflexive closure of a relation. However,
% since we can show that parallel reduction is reflexive, the transitive
% closure of parallel reduction must also be reflexive.
We say that two terms $[[a]]$
and $[[b]]$ are coherent if they can eventually reduce to some common
term $[[c]]$ through parallel reduction. The symmetric notation of
coherence suggests that it is an equivalence relation, and in fact, we
will use coherence as the equational theory for our type conversion
rule. Now, we prove that coherence is indeed an equivalence relation.

First, we show that coherence is reflexive through the following
sequence of lemmas.
\begin{lemma}[Par Refl]
  \label{lemma:parrefl}
  For all terms $[[a]]$, $[[a => a]]$.
\end{lemma}

@@

\begin{lemma}[Coherence refl]
  \label{lemma:coherencerefl}
  For all terms $[[a]]$, $[[a <=> a]]$.
\end{lemma}

Lemma~\ref{lemma:parrefl} can be proven by structural induction over
the term $[[a]]$. Lemmas~\ref{lemma:parsrefl} and \ref{lemma:coherencerefl}
immediately follow as corollaries of Lemma~\ref{lemma:parrefl}.

Symmetry of coherence immediately falls from its definition.
\begin{lemma}[Coherence sym]
  \label{lemma:coherencesym}
  If $[[a <=> b]]$, then $[[b <=> a]]$.
\end{lemma}

Before we can prove transitivity, we need to show that parallel
reduction satisfies the diamond property.
\begin{lemma}[Par cong]
  \label{lemma:parcong}
  If $[[a0 => a1]]$ and $[[b0 => b1]]$, then $[[a0 { b0 } => a1 { b1 }]]$.
\end{lemma}
\begin{lemma}[Par diamond]
  \label{lemma:pardiamond}
  If $[[a => b0]]$ and $[[a => b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 => c]]$ and $[[b1 => c]]$.
\end{lemma}
The congruence property (Lemma~\ref{lemma:parcong}) can be proven by
structural induction over the derivation of $[[a0 => a1]]$.
Likewise, Lemma~\ref{lemma:pardiamond} can be proven by structural induction
over the derivation of $[[a => b0]]$. The \rref{P-AppAbs} case requires
the use of Lemma~\ref{lemma:parcong}.

A relation that satisfies the
diamond property must also be confluent, meaning that its transitive
and reflexive closure is confluent.
\begin{lemma}[Par confluent]
  \label{lemma:parconfluent}
  If $[[a =>+ b0]]$ and $[[a =>+ b1]]$, then there exists some term
  $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$.
\end{lemma}
While $[[a =>+ b]]$ is defined as the transitive closure of $[[a => b]]$,
it coincides with the transitive and reflexive closure of $[[a => b]]$ since $[[a
=> b]]$ is reflexive (Lemma~\ref{lemma:parrefl}).

The transitivity of the coherence relation follows as a corollary of
Lemma~\ref{lemma:parconfluent}.
\begin{lemma}[Coherence trans]
  \label{lemma:coherencetrans}
  If $[[a0 <=> a1]]$ and $[[a1 <=> a2]]$, then $[[a0 <=> a2]]$.
\end{lemma}
By the definition of coherence, there exists some term $[[b]]$ such that $[[a0 =>+ b0]]$,
$[[a1 =>+ b0]]$ and some term $[[b1]]$ such that $[[a1 =>+ b1]]$ and
$[[a2 =>+ b1]]$. By Lemma~\ref{lemma:parconfluent}, there exists some
term $[[c]]$ such that $[[b0 =>+ c]]$ and $[[b1 =>+ c]]$. It sufficies
to show that $[[a0 =>+ c]]$ and $[[a2 =>+ c]]$, both of which
trivially hold since the transitive closure $[[a =>+ b]]$ is transitive.
This concludes the proof that coherence is an equivalence relation.

We note that a more conventional presentation of
\rref{T-Conv} would instead use full beta reduction as the base for
the definition of coherence. However, since full beta reduction
doesn't satisfy the diamond property, one typically needs parallel
reduction as an auxilliary definition to derive the confluence of full
beta reduction. The formulation of \lang is slightly more economical.
